### 2021-04-06 pldi21/abstract.tex
revisions up through sec 3

\begin{abstract}
    Although text editing is powerful, 
    some types of expressions are more naturally represented and manipulated
    graphically.
    % Examples include expressions that compute 
    % colors,
    % animation parameters,
    % musical sequences,
    % tabular data, 
    % plot parameters,
    % GUI widgets, 
    % and various other domain-specific data structures.
    This paper introduces \emph{live literals}, or \emph{livelits}, 
    which allow
    clients to fill holes of these types  
    by directly manipulating a     
    provider-defined GUI embedded persistently into otherwise 
    symbolic code. 
    Uniquely, livelits are compositional: the GUI can itself contain
    spliced expressions, which have full editor support and can in turn contain other livelits. 
    Splices are typed and the system 
    ensures that livelits treat splices hygienically. 
    Livelits are also uniquely live: 
    they can offer {immediate feedback} about the run-time implications 
    of the client's choices 
    even when splices mention bound variables, 
    because the system continuously gathers closures associated with the 
    hole that the livelit is tasked with filling.
    This paper introduces livelits with case studies that exercise these novel capabilities. 
    We implement livelits in Hazel, 
    a live programming environment able to 
    typecheck and run programs with holes. 
    We then define a typed lambda calculus that captures the essence of livelits as 
    live graphical macros.     
    The macro expansion process has been mechanized in Agda.
\end{abstract}

>>>>

\begin{abstract}
    Text editing is powerful, but 
    some types of expressions are more naturally represented and manipulated
    graphically.
    Examples include expressions that compute 
    colors,
    music,
    animations,
    tabular data, 
    plots,
    GUI widgets, 
    mathematical diagrams, 
    and other domain-specific data structures.
    This paper introduces \emph{live literals}, or \emph{livelits}, 
    which allow
    clients to fill holes of types like these   
    by directly manipulating a     
    user-defined GUI embedded persistently into code. 
    Uniquely, livelits are \emph{compositional}: a livelit GUI can itself embed 
    spliced expressions, which are typed, lexically scoped, enjoy full editor support and can in turn embed 
    other livelits. 
    Livelits are also uniquely \emph{live}: 
    a livelit can provide {continuous feedback} about the run-time implications 
    of the client's choices 
    even when splices mention bound variables, 
    because the system continuously gathers closures associated with the 
    hole that the livelit is tasked with filling.
    We integrate livelits into Hazel, 
    a live programming environment able to 
    typecheck and run programs with holes, 
    and describe several case studies that exercise these novel capabilities. 
    We then define a minimal typed livelit calculus, which precisely 
    specifies how livelits operate as 
    live graphical macros.     
    The metatheory of macro expansion has been mechanized in Agda.
\end{abstract}

### 2021-04-06 pldi21/case-studies.tex
revisions up through sec 3

In this section, we will detail the livelits mechanism by way of
two domain-specific case studies:
a course grade assignment case study in Sec.~\ref{sec:live-grading}
and an image transformation case study in Sec.~\ref{sec:image-transformation}.
% We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}.\todo{do we?}{}
These case studies have been implemented
in Hazel, a browser-based live programming environment for a dialect of Elm.
Elm is an industrial pure typed functional language in the
ML family used for client-side web development.
We assume basic familiarity with ML.

@@

The most common end-user application for this task is the spreadsheet, because
it allows the instructor to record grades using a natural tabular interface,
visualize this data in one of a finite number of plot styles,
and perform basic computations,
with results updated live.
However, these affordances are limited.
It is difficult to package up common operations
into reusable libraries, interact with the data using domain-specific visualizations,
and perform complex, unanticipated operations
(e.g. preparing the data in an idiosyncratic format demanded by the university's grading system).

General-purpose programming languages
can handle these scenarios, but users
lose the ability to receive live feedback and
directly manipulate data and visualizations in the editor.

@@

First, the instructor defines a value \li{grades}
that records the grades for each student using a livelit, \li{\$dataframe},
that implements a tabular user interface. The formula bar
allows the selected cell to be filled with an arbitrary Hazel expression.
For the sake of demonstration, we show a cell that has been filled using another livelit,
\li{\$slider}, in combination with symbolic manipulation.
The table itself displays not the expression itself but rather its value, just as in a spreadsheet.

Next, the instructor computes \li{averages}
for each student by applying \li{compute_averages}, a helper function
defined in a library  (not shown) shared between multiple courses.

Next, the instructor wants to ``eyeball'' reasonable \li{cutoffs} between letter grades
by directly manipulating a domain-specific livelit, \li{\$grade_cutoffs}, that provides draggable ``paddles''
superimposed on a live visualization of the distribution of \li{averages}, which is provided as a livelit parameter.
% The value of \li{cutoffs} is a labeled 4-tuple containing each cut-off.

@@

\subsection{Livelit Expansion}\label{sec:livelit-expansion}
Livelit invocations
expand to expressions.
For example, the expansion of Fig.~\ref{fig:grading} is:

\begin{lstlisting}[xleftmargin=0.2cm]
let grades = Dataframe (
  ["A1", "A2", "A3", "Midterm", "Final"],
  [("Alice", [24. +. 36. +. 33.,
              92., 83.5, 95., 88.]),
   ("Bob", [61., 64., 98., 70., 85.]),
   ("Ciri", [75., 81., 73., 82., 79.]),
   (* ... *) ]) in
let averages = compute_averages grades weights in
let cutoffs = (.A 90., .B 80., .C 70., .D 60.) in
format_for_university
  (assign_grades averages cutoffs)
\end{lstlisting}

@@

\subsection{Compositionality}\label{sec:splicing-and-parameterization}
Livelits are compositional: they can work with sub-expressions
in the form of parameters and splices.

@@

Livelit abbreviations can partially apply parameters. For example,
we can partially apply the first parameter of \li{\$slider} to define a parameterized unsigned slider livelit:
\begin{lstlisting}[numbers=none,xleftmargin=0cm]
let $uslider = $slider 0 in ...
\end{lstlisting}
Only livelits with no remaining parameters can be invoked,
so writing \li{\$uslider} in expression position will display as a ``missing livelit parameter'' error.
(In Hazel, erroneous expressions
are automatically placed inside holes and do not prevent other parts of the program from evaluating
\cite{HazelnutLive}.)
% \footnote{\label{footnote:typing}}

@@

\subsubsection{Splices}\label{sec:splices}
Spliced expressions, or \emph{splices}, appear directly inside the livelit GUI.
Splices can be filled with Hazel expressions of any form, including other livelit invocations.
For example, each cell in the \li{\$dataframe} GUI in Fig.~\ref{fig:grading}
has a corresponding splice. The formula bar at the top
allows the user to edit the splice corresponding to the selected cell,
and all of Hazel's editing affordances are available when the client does so.
% Not all splices necessarily appear in the GUI.
Unlike parameters, the number of splices can change
as the user interacts with the livelit, e.g. when changing the number of rows or columns in a \li{\$dataframe}.
% The cell itself displays the live value of the spliced expression---
% (we return to live evaluation in Sec.~\ref{sec:live-evaluation} below.

The livelit provides an expected type for each splice when it is created.
For example, the splices for the row and column keys in Fig.~\ref{fig:grading} have expected type \li{String},
and the remaining cells have expected type \li{Float}.
Hazel surfaces and uses the expected type when the cursor is in the splice \cite{hazeltutor}.
% If an expression of invalid type is entered, it will display in an error hole as usual,
% and in Hazel this will not prevent evaluation of other expressions (see Footnote \ref{footnote:typing}).

@@

\subsubsection{Hygienic Composition}\label{sec:hygiene}

Ensuring that clients can reason about binding while leaving expansions
 invisible
 requires a hygiene discipline that enforces \emph{capture avoidance}
and \emph{context independence} \cite{TLMs,adamsHygiene}.

\textbf{Capture Avoidance.}
Splices and parameters can appear anywhere in the expansion.
This becomes problematic when
they appear under a binder, e.g. in the body of a function or \li{let} binding.
Na\"ively, this could cause inadvertent capture of the bound variable by a free variable
in the parameter or splice. For example, consider a livelit that generates an expansion
of the following seemingly innocuous form:
\begin{lstlisting}[numbers=none]
let len = strlen <splice1> in
Some (<splice2> + len)
\end{lstlisting}
Here, \li{<splice2>} appears under the binding of \li{len}. If the client has filled
\li{<splice2>} with an expression that refers to a client-side binding of \li{len},
these references would na\"ively be captured. This would not occur in \li{<splice1>},
because the \li{let} is not recursive.
This breaks abstraction and is notoriously difficult to debug,
both for the livelit provider, who has no way to predict which variables a client will use,
 and the client, who does not know which variables the provider used.

To avoid this situation, parameters and splices are placed in the expansion
in a capture-avoiding manner: variables in splices
always refer to the bindings visible to the client,
rather than bindings that are hidden inside the expansion.
We discuss how this is implemented in Sec.~\ref{sec:expansion}.
% This is implemented by alpha-renaming bindings internal to the expansion as necessary.
% (We discuss potentially relaxed variations of this hygiene discipline in Sec.~\ref{sec:discussion}.)

@@

For example, there is one closure for \li{\$dataframe} in Fig.~\ref{fig:grading}.
It contains the value of \li{q1\_max} and the other variables in scope.
These values can be used to evaluate
splices that use the corresponding variables, such as the cell selected in Fig.~\ref{fig:grading}.

>>>>

In this section, we will detail the livelits mechanism by way of
two domain-specific case studies:
a course grade assignment case study in Sec.~\ref{sec:live-grading}
and an image transformation case study in Sec.~\ref{sec:image-transformation}.
% We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}.\todo{do we?}{}
These case studies have been implemented
in Hazel, a browser-based live programming environment for a 
variant of Elm \cite{Elm}.
Elm is an industrial pure typed functional language in the
ML family used for client-side web development.
We assume basic familiarity with ML.

@@

The most common end-user application for this task is the spreadsheet, because
it allows an instructor to (1) record grades using a natural tabular interface,
(2) visualize this data in one of a finite number of plot styles,
and (3) perform basic computations,
with results updated live.
However, these affordances are limited.
It is difficult to package common operations
into reusable libraries, interact with the data using domain-specific visualizations,
and perform complex, unanticipated operations
(e.g. preparing the data in an idiosyncratic format demanded by the university registrar).

General-purpose programming languages
\emph{can} handle these scenarios, but users
lose the ability to 
directly manipulate data and visualizations
and receive live (in-editor) feedback.

@@

First, the instructor defines a value \li{grades}
that records the grades for each student using a livelit, \li{\$dataframe},
that implements a tabular user interface. The formula bar
allows the selected cell to be filled with an arbitrary Hazel expression,
here a floating point expression that adds together individual problem scores,
one of which is expressed using another livelit,
\li{\$slider}.
The table itself displays not the expression itself but rather its value, \li{93.}, just as in a spreadsheet.

Next, the instructor computes \li{averages}
for each student by applying \li{compute_weighted_averages}, a helper function
defined in a library  (not shown) shared between courses.

Next, the instructor wants to ``eyeball'' reasonable letter grade \li{cutoffs} 
by directly manipulating a domain-specific livelit, \li{\$grade_cutoffs}, that provides draggable ``paddles''
superimposed on a live visualization of the distribution of \li{averages}, 
which is provided as a livelit parameter.
The value of \li{cutoffs} is a labeled 4-tuple containing each cut-off.

@@

\subsection{Livelit Expansion}\label{sec:livelit-expansion}
Livelit invocations are expressions 
that are given meaning by expansion.
For example, the expansion of Fig.~\ref{fig:grading} is:

\begin{lstlisting}[xleftmargin=0.2cm]
let grades = Dataframe (
  ["A1", "A2", "A3", "Midterm", "Final"],
  [("Alice", [24. +. 36. +. 33.,
              92., 83.5, 95., 88.]),
   ("Bob", [61., 64., 98., 70., 85.]),
   ("Carol", [75., 81., 73., 82., 79.]),
   (* ... *) ]) in
let averages = compute_averages grades weights in
let cutoffs = (.A 90., .B 80., .C 70., .D 60.) in
format_for_university
  (assign_grades averages cutoffs)
\end{lstlisting}

@@

\subsection{Compositionality}\label{sec:splicing-and-parameterization}
Livelit invocations are compositional: they can embed sub-expressions
in the form of parameters and splices.

@@

Livelit abbreviations can partially apply parameters. For example,
we can partially apply the first parameter of \li{\$slider} to define a parameterized unsigned slider livelit:
\begin{lstlisting}[numbers=none,xleftmargin=0cm]
  let $uslider = $slider 0 in ...
\end{lstlisting}
Only livelits with no remaining parameters can be invoked,
so writing \li{\$uslider} in expression position will display as a ``missing livelit parameter'' error.
In Hazel, erroneous expressions
are automatically placed inside holes and do not prevent other parts of the program from evaluating
\cite{HazelnutLive}.
A livelit invocation can also indicate that no 
expansion is available with a custom error message, e.g. due to non-sensical bounds,
which is also displayed to the user (not shown).
% \footnote{\label{footnote:typing}}

@@

\subsubsection{Splices}\label{sec:splices}
Spliced sub-expressions, or \emph{splices}, appear directly inside the livelit GUI.
Splices can be filled with Hazel expressions of any form, including other livelit invocations.
For example, each cell in the \li{\$dataframe} GUI in Fig.~\ref{fig:grading}
has a corresponding splice. The formula bar at the top
allows the user to edit the splice corresponding to the selected cell,
and all of Hazel's editing affordances are available when the client does so.
% Not all splices necessarily appear in the GUI.
Unlike parameters, the number of splices can change
as the user interacts with the livelit, e.g. when adding or removing rows or columns in a \li{\$dataframe}.
% The cell itself displays the live value of the spliced expression---
% (we return to live evaluation in Sec.~\ref{sec:live-evaluation} below.

The livelit provides an expected type for each splice when it is created.
For example, the splices for the row and column keys in Fig.~\ref{fig:grading} have expected type \li{String},
and the remaining cells have expected type \li{Float}.
Hazel displays and uses the expected type when the cursor is on the splice \cite{hazeltutor}.
% If an expression of invalid type is entered, it will display in an error hole as usual,
% and in Hazel this will not prevent evaluation of other expressions (see Footnote \ref{footnote:typing}).

@@

\subsubsection{Binding Discipline}\label{sec:hygiene}

Ensuring that clients can reason about binding while leaving expansions
 abstract
 requires enforcing two key properties: \emph{capture avoidance}
and \emph{context independence} \cite{TLMs,adamsHygiene,DBLP:conf/popl/ClingerR91}.

\textbf{Capture Avoidance.}
Splices and parameters are passed into the expansion as function arguments, 
so they cannot capture any bindings internal to the expansion. All bindings 
are lexically scoped to the livelit invocation site.

Consider an alternative where 
splices could directly appear anywhere within the expansion, 
e.g. in the body of a generated function or \li{let} binding.
Na\"ively, this could cause inadvertent capture of the bound variables in the expansion by a free variable
in the parameter or splice. For example, consider a livelit that generates an expansion
of the following seemingly innocuous form:
\begin{lstlisting}[numbers=none]
  let len = strlen <splice1> in
  if len > 0 then Some (<splice2> + len) else None 
\end{lstlisting}
Here, \li{<splice2>} appears under the binding of \li{len}. If the client has filled
\li{<splice2>} with an expression that refers to a client-side binding of \li{len},
these references would na\"ively be captured. This would not occur in \li{<splice1>},
because the \li{let} is not recursive.
This breaks abstraction and is notoriously difficult to debug,
both for the livelit provider, who has no way to predict which variables a client will use,
 and the client, who does not know which variables the provider used.

To avoid this situation, parameters and splices must be placed in the expansion
in a capture-avoiding manner: variables in splices
always refer to the bindings visible to the client,
rather than bindings that are hidden inside the expansion.
In other macro systems, e.g. that of \citet{TLMs}, 
capture avoiding substitution is used. 
% This allows splices to appear 
% multiple times and also appear in branches that may or may not evaluate. 
To simplify reasoning about live evaluation, discussed below, we take 
an even more restrictive approach by passing splices in as function 
arguments. In an eagerly evaluated language, this ensures that 
splice evaluation is not conditional on computations invisible to the 
client (such as the length check above), 
and it also ensures that evaluation will not inadvertently 
occur multiple times or not occur at all. 
% We discuss how this is implemented in Sec.~\ref{sec:expansion}.

The trade-off is that this prevents the expression of new 
binding constructs and even control flow constructs as livelits.
We take this as a strength, in that it simplifies 
reasoning for client programmers. However, this is an independent 
design decision: 
more permissive designs that nevertheless 
maintain the critical capture avoidance property are possible.

% This is implemented by alpha-renaming bindings internal to the expansion as necessary.
% (We discuss potentially relaxed variations of this hygiene discipline in Sec.~\ref{sec:discussion}.)

@@

For example, there is one closure for \li{\$dataframe} in Fig.~\ref{fig:grading}.
It contains the value of \li{q1\_max} and the other variables in scope.
These values can be used to evaluate
splices that use these variables, such as the cell selected in Fig.~\ref{fig:grading}.
